---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The data is stored in the zip file repdata-data-activity.zip. We will first, unzip this file and load the data into data frame mydt. We see that the data frame has 3 columns
- date
- interval
- steps
The date column is stored as a factor. So, we will create a new column called Date where we will convert this into POSIXct date format and get rid of the existing date column. Then we will reorder the columns. To do this, we will use the data.table, lubridate, dplyr packages

```{r LoadLibraries, results='hide', message=FALSE}
library(dplyr)
library(data.table)
library(lubridate)
```

Now, we read and preprocess the data
```{r LoadPreprocess}
unzip("./repdata-data-activity.zip")
mydt <- read.csv("activity.csv")
#convert mydt to data.table
mydt <- data.table(mydt)
mydt$Date <- ymd(mydt$date)
mydt <- mydt %>% select(-date) %>% setcolorder(c(3,1,2))
head(mydt)
```



## What is mean total number of steps taken per day?
Now, we do a histogram of the total steps taken in a day and in this histogram we will also indicate the mean and median of the total steps taken in a day (Mean will be drawn a red dotted line and Medial will be drawn as a blue dotted line). To enable us to do this histogram, we will group_by and summarize the data to get the total steps on each day. We will use the ggplot2 to draw the histogram.

```{r meanTotalSteps, message=FALSE}
t <- mydt %>% group_by(Date) %>% summarize(total_steps=sum(steps, na.rm=TRUE))
head(t)
```

Now that we have created the new data structure with the total_steps on each day, let us draw a histogram using ggplot. The mean value of total steps will be represented by a red line and the median value will be represented by a blue line.

```{r drawHist, message=FALSE}
library(ggplot2)
g <- ggplot(t, aes(x=total_steps)) + labs(title="Histogram of Total Steps") +
    ylab("frequency") +
    geom_histogram(binwidth=5000, color="black", fill="white") +
    geom_vline(aes(xintercept=mean(total_steps, na.rm=TRUE)), 
               color="red", linetype="dashed", size=1) +
    geom_vline(aes(xintercept=median(total_steps, na.rm=TRUE)), 
               color="blue", linetype="dashed", size=1)
g
```


## What is the average daily activity pattern?
To get the average daily patter, we will first create a data frame that has the average steps in every 5 minute interval across all days. Then we use this data frame to draw a plot. On the plot, we draw a line on the interval containing the maximum average steps.

```{r avg_daily_pattern}
t <- mydt %>% group_by(interval) %>% summarize(avg_steps=mean(steps, na.rm=T))
p <- ggplot(data=t, aes(x=interval, y=avg_steps)) + geom_line() +
    labs(title="Line Graph of average steps in 5 Min Intervals")
a <- t[avg_steps==max(t$avg_steps), ]
p <- p + geom_vline(aes(xintercept=a$interval), linetype="dashed")
p <- p + scale_x_continuous(breaks= c(seq(0,2000,by=500),a$interval),
                            labels=c(seq(0,2000,by=500),a$interval))
p
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
